P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations

### 2025-05-08 sage_latex_template/manuscript.tex
Rollback title
Signed-off-by: Rafael Ravedutti Lucio Machado <rafael.r.ravedutti@fau.de>
review introduction and related work
Signed-off-by: Rafael Ravedutti Lucio Machado <rafael.r.ravedutti@fau.de>

\title{P4IRS: A Code Generation Framework for Parallel and Performance-Portable Particle Simulations}
%\title{P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations}

@@

The computation of particle-pair interactions is widely used in physics simulations, especially in the fields of Molecular Dynamics and Discrete Element Method.
To efficiently compute the inter-particle contributions, different strategies can be used to increase efficiency and make use of the parallelism offered by the target hardware.
Most simulation packages, however, have their own hard-coded implementations, which are strictly tightened to their data structures and programming artifacts.
Also, several kernel implementations for the same potential or force field are present in a single package to account for different hardwares and optimizations strategies, leading to higher maintainability costs and higher portability effort.

@@

P4IRS can generate several kernel flavours of the same potential such as Linked Cells (\cite{linkedcells}) and Verlet Lists (\cite{verletlists}), where data layouts for properties and specific arrays (such as the neighbor-lists) can be switched to the case with best performance.
A runtime interface in C++ is also available and can be used to integrate P4IRS with other simulation tools, which is particularly important for integrating P4IRS simulations with our multi-physics framework waLBerLa (\cite{walberla1,walberla2}).
%as an example we show how P4IRS can use the load-balancing communication algorithm from waLBerLa to compute simulations with a more heterogeneous distribution of particles.

Most of P4IRS internal routines such as building Linked Cells and Verlet Lists are implemented in Python by constructing the intermediate representation for them with the assistance of operator overloading (see \nameref{sec:backend} section).
In this way, we leverage P4IRS analysis and transformations to achieve performance-portability for these routines as well, without requiring multiple tailored implementations.

The goals of this paper are summarized as follows:
\begin{itemize}
    \item We present our code generator tool P4IRS, describing how it works and how it can be used to generate efficient particle simulation codes.
    \item We discuss the advantages of using P4IRS to achieve performance-portability, also comparing it with other previous approaches such as tinyMD (\cite{tinymd}) and MESA-PD (\cite{mesapd1,mesapd2}).
    \item We describe P4IRS interfaces to allow its generated code to be integrated with other simulation tools, specifically in the context of domain-partitioning and communication.
    \item We provide performance results for Intel Ice Lake and AMD Milan CPUs, as well as for NVIDIA A40 and A100 accelerators to demonstrate the state-of-the-art performance of the applications generated by P4IRS in comparison to MD-Bench and MESA-PD.
    \item We display weak-scaling results in the Fritz and JUWELS Booster clusters to demonstrate the weak-scaling capabilities of P4IRS in MD and DEM simulations on modern supercomputers.
\end{itemize}

This paper is structured as follows: In the \nameref{sec:related_work} section we list related work for MD and DEM simulation packages, code generation tools for particle simulations and performance evaluation of such simulations.
In the \nameref{sec:background} section we explain the basic theory for particle simulations focusing on MD and DEM methods, also covering optimization strategies and choices.
In the \nameref{sec:pairs} section we present P4IRS with its current features and application design.
In the \nameref{sec:evaluation} section we evaluate the performance and scalability from P4IRS applications using \ac{MD} and \ac{DEM} experiments.
Finally, the \nameref{sec:conclusion} section presents our conclusion and outlook for P4IRS, also covering our future work.

@@

Diverse simulation fields can be described by updating particle trajectories based on their interactions with other particles.
In this paper, we mainly focus on \ac{MD} and \ac{DEM}, where particles represent either point masses (i.e. atoms) or rigid bodies with specific shapes (i.e. spheres, half-spaces, ellipses), respectively.
Most packages available use similar techniques to compute trajectories, but each of them have their own specific algorithms, parallelism strategies and hand-tuned implementations for specific hardware.
With P4IRS, we focus on using a single description of the problem, and then leverage its implemented routines and code transformations to produce the backend implementation for the specific target, which currently can be C++ with OpenMP and CUDA, also supporting MPI for distributed-memory parallel clusters.

In MD, there exist many packages with distinct strategies and simulation fields.
LAMMPS (\cite{lammps1,lammps2}) is a molecular package for the material-modeling field which uses the standard Verlet Lists algorithm to optimize the non-bonded short-range force calculations.
It offers many inter-atomic potentials, each one with different kernel implementations for a specific target such as CPU, GPU or Kokkos (\cite{kokkos}).
GROMACS (\cite{gromacs1,gromacs2}) is another package for MD with focus on bio-sciences simulation which uses its own Cluster Pair algorithm to optimize the short-range forces computation.
It does not support as many potentials as LAMMPS, but its algorithm tends to be significantly faster than the Verlet Lists since it employs SIMD parallelism more efficiently.
Furthermore, the optimal GROMACS kernel implementations rely on SIMD intrinsics to attain higher performance, which makes them harder to be maintained and ported to other architectures.
Also, the GPU algorithm in GROMACS differs from the CPU, since it relies on a strategy named ``super-clustering'' to reduce the amount of redundant computations caused by adjusting the cluster sizes to fit the SIMD width in the GPU.
Other MD packages such as HOOMD-blue (\cite{hoomdblue}), NAMD (\cite{namd}), AMBER (\cite{amber1}) and CHARMM (\cite{charmm}) also contain dedicated and hand-tuned implementations in general purpose languages for each supported backend.
Each version is tested, maintained, debugged and optimized independently.

@@

Solutions based on code generation are also available for specific fields of particle simulations, and these aim at providing a higher level of flexibility and extensibility while strive for efficient performance in modern accelerators.

OpenMM (\cite{openmm}) is a \ac{MD} package that makes use of Python as a front-end for simulation protocols and file I/O, it contains a C++ API for forces and integrators as well as optimized kernel implementations for CUDA, C++ and OpenCL.
Integrators and forces can be implemented by using mathematical descriptions in OpenMM, but its use cases are restricted to \ac{MD} simulations and several features required for a DEM simulation or other types of particles simulation (such as contact history and different shapes) are not available.

@@

In previous work we developed tinyMD, a proxy-app (also based on miniMD) created to evaluate the portability of MD and DEM applications with the AnyDSL (\cite{anydsl1,anydsl2}) framework.
TinyMD uses higher-order functions to abstract device iteration loops, data layouts and communication strategies, providing a domain-specific library to implement pair-wise interaction kernels that execute efficiently on multi-CPU and multi-GPU targets.
One of tinyMD's limitation is that its flexibility and optimization settings are restricted to what can be achieve through higher-order functions and the AnyDSL compilation framework which restricts the backend to the LLVM toolchain (since it generates LLVM IR at the end).
This means that the compiler optimizations from GCC and ICC cannot be applied to AnyDSL implementations (in contrast to C/C++ implementations), and one of the reasons we develop P4IRS is to overcome this restriction.

@@

Most performance-oriented researches for \ac{MD} systems use implemented packages as a black-box and adjust different simulation parameters to analyse their implications in the overall performance.
In some cases, however, in-depth performance studies were made to gain a broader understanding on efficient ways to achieve SIMD vectorization with SSE/AVX instructions (\cite{mdsimd}).
Most in-depth studies are performed using mini-apps or proxy-apps, one example is miniMD, a proxy-app that provides a simple copper face-cubic-centered lattice simulation where atom forces are computed with kernels extracted from LAMMPS for either the \ac{LJ} or \ac{EAM} potentials.

The Exascale Computing Project (ECP) Co-design Center for Particle Applications (CoPA) (\cite{ecpcopa}) focus on co-designing particle applications to prepare them for exascale computing.
They provide several proxy-apps for particle simulations such as CabanaMD and ExaMiniMD, which are designed in a modular fashion to account for different optimization strategies and particle data layouts, and also are built on top of Kokkos to be able to execute on many-core CPU and GPU targets.

@@

Based on our searches, there are not many in-depth performance studies focusing on SIMD and in-core execution for kernels used in DEM simulations, and most existing studies focus on evaluating their distributed-memory parallelism in weak-scaling settings.
With P4IRS, we intend to bridge the gap between performance strategies across different simulation domains, and use the most efficient algorithm whenever it is possible without requiring significant efforts such as re-implementing kernels and re-designing applications with tailored data structures for improving performance.

To evaluate performance in a single CPU core or GPU in this paper, we use MD-Bench and miniMD as the baseline performance, as they are well-known implementations that reach state-of-the-art performance for MD simulations.
For DEM cases, we use MESA-PD as the comparison target, since our focus is to show that we can achieve performance-portability with the flexibility that MESA-PD provides.

>>>>

%\title{P4IRS: A Code Generation Framework for Parallel and Performance-Portable Particle Simulations}
\title{P4IRS: An Intermediate Representation and Compiler for Parallel and Performance-Portable Particle Simulations}

@@

The computation of particle-pair interactions is widely used in physics simulations, especially in the fields of Molecular Dynamics and Discrete Element Method.
To efficiently compute the inter-particle contributions, different strategies can be used to increase efficiency and make use of the parallelism offered by the target hardware.
Most simulation packages, however, have their own hard-coded implementations, which are strictly tightened to their data structures and programming artifacts.
Also, several kernel implementations for the same potential or force field are present in a single package to account for different hardwares and optimizations strategies, leading to higher maintainability costs and higher portability effort\RMchange{s}.

@@

P4IRS can generate several kernel flavours of the same potential such as Linked Cells (\cite{linkedcells}) and Verlet Lists (\cite{verletlists}), where data layouts for properties and specific arrays (such as the neighbor-lists) can be switched to the case with \RMchange{the} best performance.
A runtime interface in C++ is also available and can be used to integrate P4IRS with other simulation tools, which is particularly important for integrating P4IRS simulations with our multi-physics framework waLBerLa (\cite{walberla1,walberla2}).
%as an example we show how P4IRS can use the load-balancing communication algorithm from waLBerLa to compute simulations with a more heterogeneous distribution of particles.

Most of \RMchange{the} P4IRS internal routines such as building Linked Cells and Verlet Lists are implemented in Python by constructing the intermediate representation for them with the assistance of operator overloading (see \nameref{sec:backend} section).
In this way, we leverage P4IRS analysis and transformations to achieve performance portability for these routines as well, without requiring multiple tailored implementations.

The goals of this paper are summarized as follows:
\begin{itemize}
    \item We present our code generator tool P4IRS, describing how it works and how it can be used to generate efficient particle simulation codes.
    \item We discuss the advantages of using P4IRS to achieve performance portability, also comparing it with other previous approaches such as tinyMD (\cite{tinymd}) and MESA-PD (\cite{mesapd1,mesapd2}).
    \item We describe P4IRS interfaces to allow its generated code to be integrated with other simulation tools, specifically in the context of domain-partitioning and communication.
    \item We provide performance results for Intel Ice Lake and AMD Milan CPUs, as well as for NVIDIA A40 and A100 accelerators to demonstrate the state-of-the-art performance of the applications generated by P4IRS in comparison to MD-Bench and MESA-PD.
    \item We display weak-scaling results in the Fritz and JUWELS Booster clusters to demonstrate the weak-scaling capabilities of P4IRS in MD and DEM simulations on modern supercomputers.
\end{itemize}

This paper is structured as follows: In the \nameref{sec:related_work} section we list related work for MD and DEM simulation packages, code generation tools for particle simulations and performance evaluation of such simulations.
In the \nameref{sec:background} section we explain the basic theory for particle simulations focusing on MD and DEM methods, also covering optimization strategies and choices.
In the \nameref{sec:pairs} section we present P4IRS with its current features and application design.
In the \nameref{sec:evaluation} section we evaluate the performance and scalability \RMchange{of} P4IRS applications using \ac{MD} and \ac{DEM} experiments.
Finally, the \nameref{sec:conclusion} section presents our conclusion\RMchange{s} and outlook for P4IRS, also covering our future work.

@@

Diverse simulation fields can be described by updating particle trajectories based on their interactions with other particles.
In this paper, we focus mainly on \ac{MD} and \ac{DEM}, where particles represent either point masses (i.e. atoms) or rigid bodies with specific shapes (i.e. spheres, half-spaces, ellipses), respectively.
Most available packages use similar techniques to compute trajectories, but each \RMchange{has its} own specific algorithms, parallelism strategies and hand-tuned implementations for specific hardware.
With P4IRS, we focus on using a single description of the problem, and then leverage its implemented routines and code transformations to produce the backend implementation for the specific target, which currently can be C++ with OpenMP and CUDA, also supporting MPI for distributed-memory parallel clusters.

In MD, there exist many packages with distinct strategies and simulation fields.
LAMMPS (\cite{lammps1,lammps2}) is a molecular package for the material-modeling field which uses the standard Verlet Lists algorithm to optimize the non-bonded short-range force calculations.
It offers many inter-atomic potentials, each one with different kernel implementations for a specific target such as CPU, GPU or Kokkos (\cite{kokkos}).
GROMACS (\cite{gromacs1,gromacs2}) is another package for MD with \RMchange{a} focus on \RMchange{life sciences simulations,} which uses its own Cluster Pair algorithm to optimize the \RMchange{computation of} short-range forces.
It does not support as many potentials as LAMMPS, but its algorithm tends to be significantly faster than the Verlet Lists since it employs SIMD parallelism more efficiently.
Furthermore, the optimal GROMACS kernel implementations rely on SIMD intrinsics to attain higher performance, which makes them harder to be maintained and ported to other architectures.
Also, the GPU algorithm in GROMACS differs from the CPU, since it relies on a strategy named ``super-clustering'' to reduce the amount of redundant computations caused by adjusting the cluster sizes to fit the SIMD width in the GPU.
Other MD packages such as HOOMD-blue (\cite{hoomdblue}), NAMD (\cite{namd}), AMBER (\cite{amber1}) and CHARMM (\cite{charmm}) also contain dedicated and hand-tuned implementations in general purpose languages for each supported backend.
Each version is tested, maintained, debugged and optimized independently.

@@

Solutions based on code generation are also available for specific fields of particle simulations, and these aim at providing a higher level of flexibility and extensibility while striv\RMchange{ing} for efficient performance in modern accelerators.

OpenMM (\cite{openmm}) is a \ac{MD} package that makes use of Python as a front-end for simulation protocols and file I/O, it contains a C++ API for forces and integrators as well as optimized kernel implementations for CUDA, C++ and OpenCL.
Integrators and forces can be implemented by using mathematical descriptions in OpenMM, but its use cases are restricted to \ac{MD} simulations and several features required for a DEM simulation or other types of particle simulations (such as contact history and different shapes) are not available.

@@

In previous work we developed tinyMD, a proxy-app (also based on miniMD) created to evaluate the portability of MD and DEM applications with the AnyDSL (\cite{anydsl1,anydsl2}) framework.
TinyMD uses higher-order functions to abstract device iteration loops, data layouts and communication strategies, providing a domain-specific library to implement pair-wise interaction kernels that execute efficiently on multi-CPU and multi-GPU targets.
One of \RMchange{the limitations of tinyMD} is that its flexibility and optimization settings are restricted to what can be achieved through higher-order functions and the AnyDSL compilation framework, which restricts the backend to the LLVM toolchain (since it generates LLVM IR at the end).
This means that the compiler optimizations from GCC and ICC cannot be applied to AnyDSL implementations (in contrast to C/C++ implementations), and one of the reasons we develop P4IRS is to overcome this restriction.

@@

Most performance-oriented research for \ac{MD} systems uses implemented packages as a black-box and adjust different simulation parameters to analyse their implications in the overall performance.
In some cases, however, in-depth performance studies were made to gain a broader understanding of efficient ways to achieve SIMD vectorization with SSE/AVX instructions (\cite{mdsimd}).
Most in-depth studies are performed using mini-apps or proxy-apps, one example is miniMD, a proxy-app that provides a simple copper face-cubic-centered lattice simulation where atom forces are computed with kernels extracted from LAMMPS for either the \ac{LJ} or \ac{EAM} potentials.

The Exascale Computing Project (ECP) Co-design Center for Particle Applications (CoPA) (\cite{ecpcopa}) focuses on co-designing particle applications to prepare them for exascale computing.
They provide several proxy-apps for particle simulations such as CabanaMD and ExaMiniMD, which are designed in a modular fashion to account for different optimization strategies and particle data layouts, and are built on top of Kokkos to be able to execute on many-core CPU and GPU targets.

@@

Based on our searches, there are not many in-depth performance studies focusing on SIMD and in-core execution for kernels used in DEM simulations, and most existing studies focus on evaluating their distributed-memory parallelism in weak-scaling settings.
With P4IRS, we intend to bridge the gap between performance strategies across different simulation domains, and use the most efficient algorithm whenever it is possible without requiring significant efforts such as reimplementing kernels and redesigning applications with tailored data structures \RMchange{to improve} the performance.

To evaluate performance on a single CPU core or GPU in this paper, we use MD-Bench and miniMD as the baseline performance, as they are well-known implementations that reach state-of-the-art performance for MD simulations.
For DEM cases, we use MESA-PD as the comparison target, since our focus is to show that we can achieve performance-portability with the flexibility that MESA-PD provides.
